---
title: "Data manipulation with `dplyr`"
output: github_document
---

Once you've imported data, you're going to need to do some cleaning up. 

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

```{r}
library(tidyverse)

options(tibble.print_min = 3)

litters_data = 
  read_csv("data/FAS_litters.csv")

litters_data = 
janitor::clean_names(litters_data)

pups_data = read_csv("data/FAS_pups.csv")
janitor::clean_names(pups_data)
```

### `select`

When you only need a subset of columns in a data tble; extracting only what you need can helpfully de-clutter.

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)

select(litters_data, group:gd_of_birth)
```

You can also specify columns you'd like to remove:

```{r}
select(litters_data, -pups_survive)
select(litters_data, -pups_survive, -group)
```

You can rename variables as part of this process:

```{r}
select(litters_data, GROUP=group, LiTtEr_NuMbEr = litter_number)
```

If all you want to do is rename something, you can use `rename` instead of `select`. This will rename the variables you care about, and keep everything else:

```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for `select`; read about all of them using `?select_helpers`. I use `starts_with()`, `ends_with()`, and `contains()` often, especially when there are variable names with suffixes or other standard patterns:

```{r}
select(litters_data, starts_with("gd"))
select(litters_data, ends_with("weight"))
```

I also frequently use `everything()`, which is handy for reorganizing columns without discarding anything:

```{r}
select(litters_data, litter_number, pups_survive, everything())
```

You will often filter using comparison operators (`>`, `>=`, `<`, `<=`, `==`, and `!=`). You may also use `%in%` to detect if values appear in a set (useful for characters), and `is.na()` to find missing values. The results of comparisons are logical-- the statement is `TRUE` or `FALSE` depending on the values you compare -- and can be combined with other comparisons using the logical operators `&` and `|`, or negated using `!`. 

Some ways you might filter the litters data are:

* `gd_of_birth == 20`
* `pups_born_alive >= 2`
* `pups_survive != 4`
* `!(pups_survive == 4)`
* `group %in% c("Con7", "Con8")`
* `group == "Con7" & gd_of_birth == 20`

```{r}
filter(litters_data, gd_of_birth == 20)

filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

A very common filtering step requires you to omit missing observations. You *can* do this with `filter`, but I recommend using `drop_na` from the `tidyr` package:

* `drop_na(litters_data)` will remove any row with a missing value 
* `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing. 

### `mutate`

The example below creates a new variable measuring the difference between `gd18_weight` and `gd0_weight` and modifies the existing `group` variable.

```{r}
#litter_data2 = 
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group),
       # wt_gain_kg = wt_gain * 2.2
       )
```
### `arrange`

In comparison to the preceding, arranging is pretty straightforward. You can arrange the rows in your data according to the values in one or more columns:

```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

You can also sort in descending order if you'd like.
```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)
```

### `%>%`
Piping! Control + Shift + M is the keyboard shortcut.

```{r}
litters_data = 
  read_csv("data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```


